---
title: "Comparison of the shapes of the likelihood"
author: "Jaromil Frossard and Olivier Renaud"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison of the shapes of the likelihood}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

In this tutorial, we compare the shapes of the likelihood using `lmer()` and `gANOVA()` on an example.

We need the follwing packages:

```{r, message=FALSE,warning=FALSE}
library(lme4)
library(gANOVA)
library(tidyverse)
```

# Simulating data

We can simuluate data from a model with 12 participants and a factor with 4 levels (and 3 replications per levels per participants):

```{r}
set.seed(43)

df <- expand.grid(pt = paste0("P",str_pad(1:12,width = 2,pad = "0")),
                  A = paste0("A",1:4),
                  r = 1:3,
                  stringsAsFactors = F)%>%
  select(-r)
```


```{r}
# add +(1|pt) random intercept
df<-
  df%>%
  nest(data=-pt)%>%
  mutate(ranef_pt = rnorm(n(), sd = 1.6))%>%
  unnest(data)

# add +(1|pt:A) random interaction
df<-
  df%>%
  nest(data=-c(pt,A))%>%
  mutate(ranef_ptA = rnorm(n(), sd = 3))%>%
  unnest(data)

# add an error term
df<-
  df%>%
  mutate(err = rnorm(n(),sd = 4))


# create the dependant variables
df <- df%>%
  transmute(pt = pt, A= A,
            y =  ranef_pt +ranef_ptA + err)
```


# Estimation of the model

We use the maximum likelihood estimation for both estimation:

```{r}
m_lmer <- lmer(y ~ A+ (1|pt) + (1|pt:A), data=df, contrasts = list(A = contr.sum), REML = F)
m_gANOVA <- gANOVA(y~ A + (1|pt|A), data=df, contrasts = list(A = contr.sum), REML = F)
```

We store the estimated parameters. There are 2 $\theta_i$'s associated to the random intercept and to the random interaction:

```{r}
df_opt = tibble(
  "ranef" = c("ptA", "pt"),
  "lmer" = getME(m_lmer,"theta"),
  "ganova" = getME(m_gANOVA,"theta"))%>%
  pivot_longer(c(lmer, ganova),names_to = "model",values_to = "xy")%>%
  pivot_wider(names_from = "ranef",values_from = xy)
```

Then we evaluate the optimal function for values of the parameters from 0 to 1.5, and compute the likelihood:

```{r}
df_ll <- expand_grid(
  pt = seq(from =0, to =1.5, by = 0.05),
  ptA = seq(from =0, to =1.5, by = 0.05))%>%
  mutate(lmer = map2_dbl(pt,ptA,function(x,y){getME(m_lmer,"devfun")(c(y,x))}),
         ganova = map2_dbl(pt,ptA,function(x,y){getME(m_gANOVA,"devfun")(c(y,x))}))%>%
  pivot_longer(c(lmer,ganova), names_to = "model", values_to = "log-L")%>%
  mutate(`log-L` = `log-L` *(-1/2))
```


We can compute the conditional optimum to highligh the ridge lines of the likelihood:

```{r}
df_h<- 
  df_ll%>%
  group_by(pt,model)%>%
  filter(rank(-`log-L`, ties.method="first")==1)%>%
  ungroup()%>%
  arrange(pt)

df_v<- 
  df_ll%>%
  group_by(ptA,model)%>%
  filter(rank(-`log-L`, ties.method="first")==1)%>%
  ungroup()%>%
  arrange(ptA)

```


Finally, we create a `contour plot` of the likelihood. In this case, the 2 parametrizations have the same likelihood at the optimal and the inference is the same. However using the `gANOVA` parametrization we have parameters almost orthogonal at the maximum. This behaviour comes from the orthogonalization of the random effects in `gANOVA()` and may produce an easier optimization.



```{r}
df_ll%>%
  ggplot()+
  geom_contour(aes(x = pt,y = ptA,z = `log-L`),bins = 18) +
  geom_point(data = df_opt,aes(x = pt,y = ptA),color="red",size = 3) +
  geom_path(data = df_h,aes(x = pt,y = ptA)) +
  geom_path(data = df_v,aes(x = pt,y = ptA)) +
  coord_fixed() +
  facet_grid(~model) + theme_bw()
```



















