---
title: "Spherical distribution: example"
author: "Jaromil Frossard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spherical-distribution-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

We present here some example to highlight the difference between the `gANOVA()` function and the built-in spherical distribution of `lmer()`.

We will need the following packages:

```{r,message=FALSE,warning=FALSE}
library(lme4)
library(gANOVA)
library(tidyverse)
```


Then we construct the design of a simple experiment. It contains 12 participants, a within-participant factor with 4 levels and 3 replications for each participant within each level. The design is simply:

```{r}
set.seed(42)

df <- expand.grid(pt = paste0("P",str_pad(1:12,width = 2,pad = "0")),
                  A = paste0("A",1:4),
                  r = 1:3,
                  stringsAsFactors = F)%>%
  select(-r)
```


Then we simulate the response variable


```{r}
# add +(1|pt) random intercept
df<-
  df%>%
  nest(data=-pt)%>%
  mutate(ranef_pt = rnorm(n(), sd = .8))%>%
  unnest(data)

# add +(1|pt:A) random slopes
df<-
  df%>%
  nest(data=-c(pt,A))%>%
  mutate(ranef_ptA = rnorm(n(), sd = 3))%>%
  unnest(data)

# add an error term
df<-
  df%>%
  mutate(err = rnorm(n(),sd = 4))


# create 2 dependant variables
df <- df%>%
  transmute(pt = pt, A= A,
            y_diff = ranef_pt+ranef_ptA+err,
            y_same =  ranef_pt*2+ranef_ptA+err)
```


Finally, we estimate the model using `y_same` as dependant variable:

```{r}
lmer_same <- lmer(y_same ~ A+ (1|pt) + (1|pt:A), data=df, contrasts = list(A = contr.sum))
summary(lmer_same)
```


and then gANOVA:


```{r}
gANOVA_same <- gANOVA(y_same ~ A + (1|pt|A), data=df, contrasts = list(A = contr.sum))
summary(gANOVA_same)
```

Using `y_same` the 2 estimations are the same (likelihood, fixed parameters, etc).

However, using `y_diff`, the variability of the random intercepts is smaller. The estimation will produce a singular fit using the parametrization of `lmer()`:

```{r}
lmer_diff <- lmer(y_diff~ A+ (1|pt)+(1|pt:A),data=df,contrasts = list(A = contr.sum))
summary(lmer_diff)
```


Estimating the model using `gANOVA()` reduce the constraint on the parameter space (of the variance of the intercept). the gANOVA parametrization produce a non-zero estimate of the variance of the random intercept. It follows that the likelihood is smaller:


```{r}
mod_gANOVA <- gANOVA(y_diff~ A+ (1|pt|A),data=df,contrasts = list(A = contr.sum))
summary(mod_gANOVA)
```





